---
title: Progress
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, include = FALSE}
# install.packages(c("rjson", "tidyverse", "lubridate", "knitr", "readxl"))
library(knitr) # formatting tables in markdown
library(rjson) # for reading api
library(tidyverse) # for visualisations
library(lubridate) # for converting character strings to dates
library(readxl) # for reading Excel spreadsheets into R
```

```{r data sources, include = FALSE}
vaxxdata <-
  (
    "https://data.ontario.ca/api/3/action/datastore_search?resource_id=8a89caa9-511c-4568-af89-7f2174b4378c&limit=1000"
  ) #API set to capture first 1000 records.

# Get population data.
ONpop <-
  (
    "https://data.ontario.ca/datastore/dump/775ca815-5028-4e9b-9dd4-6975ff1be021?bom=True"
  )
age <- read.csv(ONpop)
latestage <- tail(age, n = 10)
ONpop12plus <-
  latestage[9, 6] + latestage[1, 6] # Capture total adult 18+ and 12-17 year old population totals from the Ontario COVID-19 by age dataset. Add them together to give total vaccine eligible population.
```

```{r vaccination data, include = FALSE}
# Transform vaccination data into a data frame.
vaxx <- fromJSON(file = vaxxdata) # Read in API JSON data
vaxx <- (vaxx$result$records) # Isolate vaccination data fields
vaxx <-
  as.data.frame(do.call(cbind, vaxx)) # Bind days into a data frame.
vaxx <- t(vaxx) # transpose dataframe
vaxx <-
  apply(vaxx, 2, function(y)
    sapply(y, function(x)
      ifelse(is.null(x), 0, x))) # Replace null values with zeros.
vaxx <- as.data.frame(vaxx)
```

```{r clean vaccination data, include = FALSE}
# Clean vaccination data removing "_" from "_id".
names(vaxx)[names(vaxx) == "_id"] <- "id"

# Convert report data to date
vaxx$report_date <-
  ymd_hms(vaxx$report_date) # converts report date into date.

# Convert other columns to number
vaxx$id <- as.numeric(as.character(vaxx$id))
vaxx$previous_day_total_doses_administered <-
  as.numeric(as.character(vaxx$previous_day_total_doses_administered))
vaxx$previous_day_at_least_one <-
  as.numeric(as.character(vaxx$previous_day_at_least_one))
vaxx$previous_day_fully_vaccinated <-
  as.numeric(as.character(vaxx$previous_day_fully_vaccinated))
vaxx$total_doses_administered <-
  as.numeric(as.character(vaxx$total_doses_administered))
vaxx$total_individuals_at_least_one <-
  as.numeric(as.character(vaxx$total_individuals_at_least_one))
vaxx$total_doses_in_fully_vaccinated_individuals <-
  as.numeric(as.character(vaxx$total_doses_in_fully_vaccinated_individuals))
vaxx$total_individuals_fully_vaccinated <-
  as.numeric(as.character(vaxx$total_individuals_fully_vaccinated))
```

```{r missing christmas data, include = FALSE}
# The province administered the first vaccine on 2020-12-14 but didn't record data until 2020-12-24.
firstdoses <- head(vaxx$total_doses_administered, n = 1)
firstdose2020data <- data.frame(
  "id" = 0,
  "report_date" = c(
    "2020-12-15T00:00:00",
    "2020-12-16T00:00:00",
    "2020-12-17T00:00:00",
    "2020-12-18T00:00:00",
    "2020-12-19T00:00:00",
    "2020-12-20T00:00:00",
    "2020-12-21T00:00:00",
    "2020-12-22T00:00:00",
    "2020=12-23T00:00:00"
  ),
  "previous_day_total_doses_administered" = 0,
  "previous_day_at_least_one" = 0,
  "previous_day_fully_vaccinated" = 0,
  "total_doses_administered" = 0,
  "total_individuals_at_least_one" = 0,
  "total_doses_in_fully_vaccinated_individuals" = 0,
  "total_individuals_fully_vaccinated" = 0
)

# The province did not administer any vaccines between 2020-12-25 and 2020-12-29 but omits this data from the data set. Add missing Christmas 2020 values.
christmas2020data <- data.frame(
  "id" = 0,
  "report_date" = c(
    "2020-12-25T00:00:00",
    "2020-12-26T00:00:00",
    "2020-12-27T00:00:00",
    "2020-12-28T00:00:00",
    "2020-12-29T00:00:00"
  ),
  "previous_day_total_doses_administered" = c(firstdoses, 0, 0, 0, 0),
  "previous_day_at_least_one" = firstdoses,
  "previous_day_fully_vaccinated" = 0,
  "total_doses_administered" = firstdoses,
  "total_individuals_at_least_one" = firstdoses,
  "total_doses_in_fully_vaccinated_individuals" = 0,
  "total_individuals_fully_vaccinated" = 0
)

# converts report date into date.
firstdose2020data$report_date <- ymd_hms(firstdose2020data$report)
christmas2020data$report_date <-
  ymd_hms(christmas2020data$report_date)
```

```{r merge first doses and christmas 2020 data into data set, include = FALSE}
vaxx <-
  rbind(vaxx, firstdose2020data, christmas2020data) # merge with vaxx data
vaxx <- vaxx[order(vaxx$report_date),] # sort by report date
```

```{r create vaccination percentages, include = FALSE}
# Calculate new columns for vaccination of percentages of adults either partly vaccinated or fully vaccinated using adult population estimate.

# Total number of partially vaccinated adults
vaxx$part <-
  ((
    vaxx$total_doses_administered - vaxx$total_doses_in_fully_vaccinated_individuals
  )
  )

# Total number of adults either fully or partially vaccinated
vaxx$fullorpart <-
  ((vaxx$total_individuals_fully_vaccinated + vaxx$part))

# Number of adults who are neither fully or partially vaccinated
vaxx$none <- ((ONpop12plus - vaxx$fullorpart))

# Percentage shares
vaxx$percentagepart <-
  ((vaxx$part / ONpop12plus) * 100) # Part vaccinated
vaxx$percentagefully <-
  ((vaxx$total_individuals_fully_vaccinated / ONpop12plus) * 100) # Fully vaccinated
vaxx$percentagefullorpart <-
  ((vaxx$fullorpart / ONpop12plus) * 100) # Fully or partly vaccinated
vaxx$percentagenone <-
  ((vaxx$none / ONpop12plus) * 100) # No vaccination

# Round percentages to two decimals
vaxx$percentagefully <- round(vaxx$percentagefully, digits = 1)
vaxx$percentagepart <- round(vaxx$percentagepart, digits = 1)
vaxx$percentagefullorpart <-
  round(vaxx$percentagefullorpart, digits = 1)
vaxx$percentagenone <- round(vaxx$percentagenone, digits = 1)
```

```{r unlist vaccination data, include = FALSE}
# Subset percentage data for visualisation.
vizfullorpart <-
  subset(vaxx, select = c("report_date", "percentagefullorpart"))
vizfull <-
  subset(vaxx, select = c("report_date", "percentagefully"))
vizpart <- subset(vaxx, select = c("report_date", "percentagepart"))

# standardise column names
names(vizfullorpart)[names(vizfullorpart) == "percentagefullorpart"] <-
  "percentage"
names(vizfull)[names(vizfull) == "percentagefully"] <- "percentage"
names(vizpart)[names(vizpart) == "percentagepart"] <- "percentage"
```


### Vaccinated
Vaccination status as a percentage of the population
```{r create subset of vaccination status as percentage of adult population, include = FALSE}
bar <- tail(vaxx, n = 1)
bar <-
  subset(bar, select = c(percentagefully, percentagepart, percentagenone))
bar <- t(bar)
bar <- as_tibble(bar)
row.names(bar) <-
  c("Fully vaccinated", "Partly vaccinated", "No vaccination")
bar <- tibble::rownames_to_column(bar, "Status")
colnames(bar) <- c("Status", "Percentage")
# Round percentages to two decimals
bar$Percentage <- round(bar$Percentage, digits = 1)
```

```{r latest number of fully vaccinated, results = 'asis', echo = FALSE}
latest <-
  subset(vaxx,
         select = c(
           report_date,
           total_individuals_fully_vaccinated,
           part,
           fullorpart
         ))
colnames(latest) <-
  c("Report date", "Fully", "Partly", "Fully or partly ")
latest <- tail(latest, n = 1)
knitr::kable(latest, row.names = FALSE, align = "c")
```

```{r barplot visualisation of vaccination status as percentage of adult population, echo = FALSE}
bar$Status <-
  factor(bar$Status,
         levels = c("No vaccination", "Partly vaccinated", "Fully vaccinated"))
ggplot(bar, aes(x = "", y = Percentage, fill = Status)) +
  geom_bar(stat = "identity", width = .375) +
  geom_text(aes(label = paste0(Percentage, sep = "%")),
            position = position_stack(vjust = .5),
            size = 2.5) +
  labs(
    x = NULL,
    y = NULL,
    caption = "Data: Ontario COVID-19 vaccine data, Ontario COVID-19 vaccine data by age",
    title = "Ontario COVID-19 vaccination status of population (aged 12+)",
    subtitle = Sys.Date()
  ) +
  theme_classic() +
  theme(
    axis.line = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) +
  scale_fill_brewer(palette = "Greys") +
  coord_flip()
```


### Progress
```{r vaccine percentage population visualisation, echo = FALSE}
ggplot(vizfull, aes(report_date, percentage)) +
  geom_line(aes(linetype = "Fully")) +
  geom_line(data = vizpart, aes(linetype = "Partly")) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(
    linetype = "Vaccinated",
    x = "",
    y = "%",
    caption = "Data: Ontario COVID-19 vaccine data, Ontario COVID-19 vaccine data by age",
    title = "Ontario COVID-19 vaccination, 2020-2021",
    subtitle = "Percentage of population (aged 12+)"
  ) +
  theme(legend.position = "right", panel.grid.minor = element_blank())
```


### Vaccine doses
```{r latest number of vaccine doses, results = 'asis', echo = FALSE}
latestdoses <-
  subset(
    vaxx,
    select = c(
      report_date,
      previous_day_total_doses_administered,
      total_doses_administered
    )
  )
colnames(latestdoses) <-
  c("Report date",
    "Previous day doses administered",
    "Total number of doses administered")
latestdoses <- tail(latestdoses, n = 1)
knitr::kable(latestdoses, row.names = FALSE, align = "c")
```
  
  
### Daily and cumulative vaccine doses  
```{r daily and cumilative totals of vaccine administered, echo = FALSE}
# Subset data
dosesadministered <-
  subset(
    vaxx,
    select = c(
      "report_date",
      "previous_day_total_doses_administered",
      "total_doses_administered"
    )
  )

# Create new column for data administered using reported date minus 24 hours in seconds.
dosesadministered$administered_date <-
  ((dosesadministered$report_date - 86400))
dosesadministered <-
  subset (dosesadministered, select = -report_date) # drop reported date column
names(dosesadministered) <-
  c("doses", "totaldoses", "administered") # rename columns
dosesadministered$administered <-
  as.Date(dosesadministered$administered) # ensure date is a date field and not character.

# Visualisation
ggplot() +
  labs(title = "Ontario COVID-19 vaccination",
       subtitle = "Daily doses of vaccine and cumilative doses administered, 2020-2021",
       caption = "Source: Ontario COVID-19 vaccine data") +
  geom_bar(
    mapping = aes(x = dosesadministered$administered, y = dosesadministered$doses),
    stat = "identity"
  ) +
  geom_line(
    mapping = aes(x = dosesadministered$administered, y = dosesadministered$totaldoses *
                    .01),
    size = 1,
    color = "blue",
  ) +
  scale_x_date(name = "") +
  scale_y_continuous(name = "Daily doses",
                     sec.axis = sec_axis(~ . / 10000, name = "Cumilative (million)")) +
  theme(
    axis.title.y = element_text(color = "black"),
    axis.title.y.right = element_text(color = "blue"),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

### Age group
```{r latest age group data, echo = FALSE}
latestage <-
  subset(latestage,
         Agegroup != "Adults_18plus" &
           Agegroup != "Undisclosed_or_missing") # Remove the adult total and NA groups.
latestage <-
  mutate(latestage,
         At.least.one.dose_cumulative = At.least.one.dose_cumulative - Second_dose_cumulative) # Subtract Second dose total from at least one total.

# Add categorical variable for partly and fully vaccinated individuals.
age1 <- latestage %>% select(Agegroup, At.least.one.dose_cumulative)
age2 <- latestage %>% select(Agegroup, Second_dose_cumulative)
age1 <- rename(age1, Total = At.least.one.dose_cumulative)
age2 <- rename(age2, Total = Second_dose_cumulative)
age1$Status <- "Partly"
age2$Status <- "Fully"
age <- rbind(age1, age2)
```

```{r visualisation for vaccination by age group, echo = FALSE}
age$Status <- factor(age$Status, levels = c("Partly", "Fully"))
ggplot(data = age, aes(x = Agegroup, y = Total, fill = Status)) +
  geom_bar(stat = "identity") +
  labs(
    y = "Individuals",
    x = "",
    caption = "Data: Ontario COVID-19 vaccine data by age",
    title = "Ontario COVID-19 vaccination status of population (aged 12+)",
    subtitle = Sys.Date()
  ) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.x = element_line(colour = "grey", size = 0.25),
    panel.grid.minor.x = element_blank(),
    panel.background = element_rect(fill = "#FFFFFF", color = "white"),
  ) +
  scale_fill_brewer(palette = "Greys") +
  coord_flip()
```

### Region
```{r create region download file, include = FALSE}
# Copy link does not point to a file directly. Need to create a spreadsheet in which to download data.
file.create("ICES-COVID19-Vaccination-Data-by-FSA.xlsx")
download.file(
  "https://www.ices.on.ca/~/media/Files/COVID-19/ICES-COVID19-Vaccination-Data-by-FSA.ashx?la=en-CA",
  "ICES-COVID19-Vaccination-Data-by-FSA.xlsx"
)
```

```{r download region data, echo = FALSE}
fsa1 <-
  tibble(
    read_xlsx(
      "ICES-COVID19-Vaccination-Data-by-FSA.xlsx",
      sheet = "At least 1 Dose Vaccine by FSA",
      skip = 23,
      col_names = TRUE
    )
  )
fsa2 <-
  tibble(
    read_xlsx(
      "ICES-COVID19-Vaccination-Data-by-FSA.xlsx",
      sheet = "Both Dose Vaccination by FSA",
      skip = 23,
      col_names = TRUE
    )
  )
```

```{r rename columns in region data, echo = FALSE}
names(fsa1)[names(fsa1) == "% Vaccinated with at least 1 dose\r\n(All ages, including <12 and undocumented age)"] <-
  "onedose"
names(fsa2)[names(fsa2) == "% Vaccinated with both doses\r\n(All ages, including <12 and undocumented age)"] <-
  "bothdose"
```

```{r keep only colums wanted, echo = FALSE}
# We only want to keep the FSA region and percentage vaccinated for all ages columns.
fsa1 <- select(fsa1, FSA, onedose)
fsa2 <- select(fsa2, FSA, bothdose)
```

```{r merge data from the two worksheets into one data frame, echo = FALSE}
# Merge on the FSA value
fsavaxx <- tibble(merge(fsa1, fsa2, by.x = "FSA"))
```

```{r create region variable, echo = FALSE}
# Duplicate the FSA column but keeping only the first letter
fsavaxx$region <- gsub("([A-Za-z]+).*", "\\1", fsavaxx$FSA)

# Recode that letter into region names. Info on FSAs for Ontario: https://en.wikipedia.org/wiki/Postal_codes_in_Canada#Table_of_all_postal_codes
fsavaxx$region <- recode_factor(
  fsavaxx$region,
  K = "Eastern",
  L = "Central",
  M = "Toronto",
  N = "Western",
  P = "Northern",
)
```

```{r create subsets by region, echo = FALSE}
# Split the data frame into sub-sets based on their region.
eastern <- fsavaxx[fsavaxx$region == "Eastern", ]
central <- fsavaxx[fsavaxx$region == "Central", ]
toronto <- fsavaxx[fsavaxx$region == "Toronto", ]
western <- fsavaxx[fsavaxx$region == "Western", ]
northern <- fsavaxx[fsavaxx$region == "Northern", ]
```

```{r average percent value for each region, echo = FALSE}
# Find the mean value for percentages vaccinated with one dose and both doses for each region,
Eastern <- colMeans(eastern[, c("onedose", "bothdose")])
Central <- colMeans(central[, c("onedose", "bothdose")])
Toronto <- colMeans(toronto[, c("onedose", "bothdose")])
Western <- colMeans(western[, c("onedose", "bothdose")])
Northern <- colMeans(northern[, c("onedose", "bothdose")])
```

```{r bind region means into a tibble, echo = FALSE}
region_mean <-
  as.data.frame(rbind(Eastern, Central, Toronto, Western, Northern))
region_mean <-
  tibble::rownames_to_column(region_mean, "Region") # make sure first row is column names
```

```{r set up a data frame for visualisation, echo = FALSE}
# Break into separate data frames depending on one dose or both dose values
region_viz1 <- tibble(region_mean %>% select(Region, onedose))
region_viz2 <- tibble(region_mean %>% select(Region, bothdose))

# Standardise the column name for percentage values in those data frames
region_viz1 <- region_viz1 %>% rename(PartRegion = Region)
region_viz2 <- region_viz2 %>% rename(FullRegion = Region)

# Multiply values by 100
region_viz1$onedose <- region_viz1$onedose * 100
region_viz2$bothdose <- region_viz2$bothdose * 100

# Add a category column for both subsets depending on their states as either a one dose or both dose value.
region_viz1 <- mutate(region_viz1, PartStatus = "Partly")
region_viz2 <- mutate(region_viz2, FullStatus = "Fully")

# Bind the two subsets together as columns.
region_viz <- cbind(region_viz1, region_viz2) # bind
region_viz <-
  region_viz %>% mutate(onedose = onedose - bothdose) # Difference between full and part vaccinated
region_viz <- region_viz %>% mutate(
  NoRegion = FullRegion,
  nodose = 100 - (onedose + bothdose),
  NoStatus = "None"
) # Calculate percentage not vaccinated.

# Write back to indivisual tables.
region_viz1 <- select(region_viz, PartRegion, onedose, PartStatus)
region_viz2 <- select(region_viz, FullRegion, bothdose, FullStatus)
region_viz3 <- select(region_viz, NoRegion, nodose, NoStatus)

# Rename columns for consistency
region_viz1 <-
  rename(region_viz1,
         Region = PartRegion,
         Percent = onedose,
         Status = PartStatus)
region_viz2 <-
  rename(region_viz2,
         Region = FullRegion,
         Percent = bothdose,
         Status = FullStatus)
region_viz3 <-
  rename(region_viz3,
         Region =  NoRegion,
         Percent = nodose,
         Status = NoStatus)

# Bind together Full, Part, and No vaccination subsets.
region_viz <- tibble(rbind(region_viz1, region_viz2, region_viz3))

# Rounds percent value to one digit.
region_viz$Percent <- round(region_viz$Percent, digits = 0)

# Order them for visualisation on one dose vale first in the bar chart.
region_viz$Status <-
  factor(region_viz$Status, levels = c("None", "Partly", "Fully"))
```

```{r stacked bar chart visualisation for region, echo = FALSE}
ggplot(data = region_viz, aes(x = Region, y = Percent, fill = Status)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(Percent, sep = "%")),
            position = position_stack(vjust = .5),
            size = 2.5) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(
    y = "%",
    x = "",
    caption = "Data: ICES COVID-19 Dashboard vaccine coverage estimates for selected age groups by Forward Sortation Areas (FSAs).",
    title = "Estimated Ontario COVID-19 vaccination status of population (aged 12+)\npercentage by FSA region",
    subtitle = Sys.Date()
  ) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.x = element_line(colour = "grey", size = 0.25),
    panel.grid.minor.x = element_blank(),
    panel.background = element_rect(fill = "#FFFFFF", color = "white"),
  ) +
  scale_fill_brewer(palette = "Greys") +
  coord_flip()
```


## Citations
### Data
Ontario COVID-19 Vaccine Data https://data.ontario.ca/dataset/covid-19-vaccine-data-in-ontario/resource/8a89caa9-511c-4568-af89-7f2174b4378c?view_id=9e42f55b-723f-46dd-b0d9-643670e01fed published under Open Government Licence – Ontario version 1.0 (https://www.ontario.ca/page/open-government-licence-ontario)

Ontario COVID-19 Vaccine Data by Age https://data.ontario.ca/dataset/covid-19-vaccine-data-in-ontario/resource/775ca815-5028-4e9b-9dd4-6975ff1be021 published under Open Government Licence – Ontario version 1.0 (https://www.ontario.ca/page/open-government-licence-ontario)

ICES COVID-19 Dashboard vaccine coverage estimates for selected age groups by Forward Sortation Areas (FSAs)
https://www.ices.on.ca/~/media/Files/COVID-19/ICES-COVID19-Vaccination-Data-by-FSA.ashx?la=en-CA available at https://www.ices.on.ca/DAS/AHRQ/COVID-19-Dashboard

### Code
Horton, L. (2021). Ontario COVID-19 Vaccinations. *GitHub repository*, https://github.com/laurencehorton/ontario-covid-vaccinations published under Creative Commons Zero v1.0 Universal https://github.com/laurencehorton/ontario-covid-vaccinations/blob/main/LICENSE 

```{r citations, include = FALSE}
toBibtex(citation("base"))
toBibtex(citation("knitr"))
toBibtex(citation("lubridate"))
toBibtex(citation("rjson"))
toBibtex(citation("tidyverse"))
toBibtex(citation("rmarkdown"))
toBibtex(citation("readxl"))