---
title: Progress
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, include = FALSE}
# install.packages(c("rjson", "tidyverse", "lubridate", "knitr"))
library(knitr) # formatting tables in markdown
library(rjson) # for reading api
library(tidyverse) # for visualisations
library(lubridate) # for converting character strings to dates
```

```{r data sources, include = FALSE}
vaxxdata <-
  (
    "https://data.ontario.ca/api/3/action/datastore_search?resource_id=8a89caa9-511c-4568-af89-7f2174b4378c"
  )
ONpop2020 <-
  read.csv(
    "https://www150.statcan.gc.ca/t1/tbl1/en/dtl!downloadDbLoadingData-nonTraduit.action?pid=1710000501&latestN=0&startDate=20200101&endDate=20200101&csvLocale=en&selectedMembers=%5B%5B7%5D%2C%5B1%5D%2C%5B1%2C7%2C13%2C19%2C25%2C31%2C37%2C43%2C49%2C55%2C61%2C67%2C73%2C79%2C85%2C86%2C87%2C88%2C89%2C105%2C126%2C132%2C138%5D%5D"
  )
```

```{r population, include = FALSE}
## Clean data  
# Slice Ontario population estimates to obtain a value for population over 15.
ONpop2020 <-
  ONpop2020[c("Age.group", "VALUE")] # Keep only these columns
ONpop2020 <-
  ONpop2020[-c(1, 23), ] # Remove total population and median age rows
ONpop2020all <-
  sum(ONpop2020["VALUE"]) # obtain sum of total population as a numeric value
ONpop2020kids <-
  ONpop2020 %>% slice(1:3) # remove age categories 0-4, 5-9, 10-14 from all population.
ONpop2020kids <-
  sum(ONpop2020kids["VALUE"]) # Sum the age if categories 0-4, 5-9, 10-14.
ONpop2020adults <-
  ONpop2020all - ONpop2020kids # Subtract sum total of values 0-4, 5-9, 10-14 from total population to obtain adult population estimate.
```

```{r vaccination data, include = FALSE}
# Transform vaccination data into a data frame.
vaxx <- fromJSON(file = vaxxdata) # Read in API JSON data
vaxx <- (vaxx$result$records) # Isolate vaccination data fields
vaxx <-
  as.data.frame(do.call(cbind, vaxx)) # Bind days into a data frame.
vaxx <- t(vaxx) # transpose dataframe
vaxx <- as.data.frame(vaxx) # Save as data frame.
```

```{r clean vaccination data, include = FALSE}
# Clean vaccination data by removing commas and removing "_" from "_id".
vaxx$previous_day_doses_administered <-
  gsub(",", "", vaxx$previous_day_doses_administered)
vaxx$total_doses_administered <-
  gsub(",", "", vaxx$total_doses_administered)
vaxx$total_doses_in_fully_vaccinated_individuals <-
  gsub(",", "", vaxx$total_doses_in_fully_vaccinated_individuals)
vaxx$total_individuals_fully_vaccinated <-
  gsub(",", "", vaxx$total_individuals_fully_vaccinated)
names(vaxx)[names(vaxx) == "_id"] <- "id"
```

```{r convert vaccination data to numeric, include = FALSE}
# Convert strings in vaccination data into numbers.
vaxx$previous_day_doses_administered <-
  as.numeric(as.list(vaxx$previous_day_doses_administered))
vaxx$total_doses_administered <-
  as.numeric(as.list(vaxx$total_doses_administered))
vaxx$total_doses_in_fully_vaccinated_individuals <-
  as.numeric(as.list(vaxx$total_doses_in_fully_vaccinated_individuals))
vaxx$total_individuals_fully_vaccinated <-
  as.numeric(as.list(vaxx$total_individuals_fully_vaccinated))
vaxx$report_date <-
  ymd_hms(vaxx$report_date) # converts report date into date.
vaxx[is.na(vaxx)] = 0 # convert NAs to zero.
```

```{r missing christmas data, include = FALSE}
# The province did not administer any vaccines between 2020-12-25 and 2020-12-29 but omits this data from the data set. Add missing Christmas 2020 values.
christmas2020data <- data.frame(
  id = c(0, 0, 0, 0, 0),
  report_date = c(
    "2020-12-25T00:00:00",
    "2020-12-26T00:00:00",
    "2020-12-27T00:00:00",
    "2020-12-28T00:00:00",
    "2020-12-29T00:00:00"
  ),
  "previous_day_doses_administered" = c(10756, 0, 0, 0, 0),
  "total_doses_administered" = c(10756, 10756, 10756, 10756, 10756),
  "total_doses_in_fully_vaccinated_individuals" = c(0, 0, 0, 0, 0),
  "total_individuals_fully_vaccinated" = c(0, 0, 0, 0, 0)
)
christmas2020data$report_date <-
  ymd_hms(christmas2020data$report_date) # converts report date into date.
vaxx <- rbind(vaxx, christmas2020data) # merge with vaxx data
vaxx <- vaxx[order(vaxx$report_date),] # sort by report date
```

```{r create vaccination percentages, include = FALSE}
# Calculate new columns for vaccination of percentages of adults either partly vaccinated or fully vaccinated using adult population estimate.
vaxx$percentagepart <-
  ((vaxx$total_doses_administered / ONpop2020adults) * 100)
vaxx$percentagefully <-
  ((vaxx$total_individuals_fully_vaccinated / ONpop2020adults) * 100)
```

```{r unlist vaccination data, include = FALSE}
# Unlist the vaccination data for visualisation.
vaxx <- as.data.frame(lapply(vaxx, unlist)) # Unlist vaccination data

# Subset percentage data for visualisation.
vizpart <- subset(vaxx, select = c("report_date", "percentagepart"))
vizfull <-
  subset(vaxx, select = c("report_date", "percentagefully"))

# standardise column names
names(vizpart)[names(vizpart) == "percentagepart"] <- "percentage"
names(vizfull)[names(vizfull) == "percentagefully"] <- "percentage"
```
  
### Fully vaccinated  
```{r latest number of fully vaccinated, results = 'asis', echo = FALSE}
latest <- subset(vaxx, select = c(report_date, total_individuals_fully_vaccinated, percentagefully))
latest$percentagefully <- round(latest$percentagefully, digits = 2)
colnames(latest) <- c("Report date", "Individuals fully vaccinated", "Percentage of Ontario adult population fully vaccinated")
latest <- tail(latest, n=1)
knitr::kable(latest, row.names = FALSE, align = "c")
```
  

### Progress as a percentage of adult population  
```{r vaccine percentage population visualisation, echo = FALSE}
ggplot(vizfull, aes(report_date, percentage)) +
  geom_line(aes(linetype = "Fully")) +
  geom_line(data = vizpart, aes(linetype = "Fully or Partly")) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(
    linetype = "Vaccinated",
    x = "",
    y = "",
    caption = "Data: Ontario COVID-19 vaccine data and StatsCan Ontario population estimate 2020",
    title = "Ontario COVID-19 vaccination, 2020-2021",
    subtitle = "Percentage of adult population (15+)"
  ) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank()
  )
```
  
### Vaccine doses
```{r latest number of vaccine doses, results = 'asis', echo = FALSE}
latestdoses <- subset(vaxx, select = c(report_date, previous_day_doses_administered, total_doses_administered))
colnames(latestdoses) <- c("Report date", "Previous day doses administered", "Total number of doses administered")
latestdoses <- tail(latestdoses, n=1)
knitr::kable(latestdoses, row.names = FALSE, align = "c")
```
  
  
### Daily and cumulative vaccine doses  
```{r daily and cumilative totals of vaccine administered, echo = FALSE}
# Subset data
dosesadministered <-
  subset(
    vaxx,
    select = c(
      "report_date",
      "previous_day_doses_administered",
      "total_doses_administered"
    )
  )

# Create new column for data administered using reported date minus 24 hours in seconds.
dosesadministered$administered_date <-
  ((dosesadministered$report_date - 86400))
dosesadministered <-
  subset (dosesadministered, select = -report_date) # drop reported date column
names(dosesadministered) <-
  c("doses", "totaldoses", "administered") # rename columns
dosesadministered$administered <-
  as.Date(dosesadministered$administered) # ensure date is a date field and not character.

# Visualisation
ggplot() +
  labs(title = "Ontario COVID-19 vaccination",
       subtitle = "Daily doses of vaccine and cumilative doses administered, 2020-2021",
       caption = "Source: Ontario COVID-19 vaccine data") +
  geom_bar(
    mapping = aes(x = dosesadministered$administered, y = dosesadministered$doses),
    stat = "identity"
  ) +
  geom_line(
    mapping = aes(x = dosesadministered$administered, y = dosesadministered$totaldoses *
                    .01),
    size = 1,
    color = "blue",
  ) +
  scale_x_date(name = "") +
  scale_y_continuous(name = "Daily doses",
                     sec.axis = sec_axis(~ . / 10000, name = "Cumilative (million)")) +
  theme(
    axis.title.y = element_text(color = "black"),
    axis.title.y.right = element_text(color = "blue"),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

## Progress targets  
Projecting time required at the current rate of adults (aged 15+) becoming fully vaccinated until percentages of the adult population are fully vaccinated.

Assumptions:  
* Population based on StatsCan's 2020 projection for Ontario minus share of the estimated under the age of 15.  
* Average rate at which people become fully vaccinated is based on latest reported figure.  
* Assumes vaccinations continue every day of the year.  

```{r subset individuals fully vaccinated, include = FALSE}
# Subset data on total individuals fully vaccinated.
individualsfullyvaccinated <-
  subset(
    vaxx,
    select = c(
      "report_date",
      "total_individuals_fully_vaccinated"
    )
  )
```

```{r change in total of adults fully vaccinated, include = FALSE}
# Calculate the daily reported change in the number of adults who become fully vaccinated.
individualsfullyvaccinated <- individualsfullyvaccinated %>%
    mutate(change = total_individuals_fully_vaccinated - lag(total_individuals_fully_vaccinated, default = first(total_individuals_fully_vaccinated)))
```

```{r calculate percentages of projected adult population, include = FALSE}
# Calculate values for the percentages of the estimated Ontario adult population
ONpop2020adult95 <- ONpop2020adults / 100 * 95
ONpop2020adult90 <- ONpop2020adults / 100 * 90
ONpop2020adult85 <- ONpop2020adults / 100 * 85
ONpop2020adult80 <- ONpop2020adults / 100 * 80
ONpop2020adult75 <- ONpop2020adults / 100 * 75
ONpop2020adult70 <- ONpop2020adults / 100 * 70
ONpop2020adult65 <- ONpop2020adults / 100 * 65
ONpop2020adult60 <- ONpop2020adults / 100 * 60
ONpop2020adult55 <- ONpop2020adults / 100 * 55
ONpop2020adult50 <- ONpop2020adults / 100 * 50
ONpop2020adult45 <- ONpop2020adults / 100 * 45
ONpop2020adult40 <- ONpop2020adults / 100 * 40
ONpop2020adult35 <- ONpop2020adults / 100 * 35
ONpop2020adult30 <- ONpop2020adults / 100 * 30
ONpop2020adult25 <- ONpop2020adults / 100 * 25
ONpop2020adult20 <- ONpop2020adults / 100 * 20
ONpop2020adult15 <- ONpop2020adults / 100 * 15
ONpop2020adult10 <- ONpop2020adults / 100 * 10
ONpop2020adult05 <- ONpop2020adults / 100 * 05
```

```{r latest reported population fully vaccinated total, include = FALSE}
latestfullvaxx <- tail(vaxx$total_individuals_fully_vaccinated, n = 1)
```

```{r remaining population to be vaccinated, include = FALSE}
# Subtract number of people fully vaccinated from population targets
target100 <- (ONpop2020adults - latestfullvaxx)
target95 <- (ONpop2020adult95 - latestfullvaxx)
target90 <- (ONpop2020adult90 - latestfullvaxx)
target85 <- (ONpop2020adult85 - latestfullvaxx)
target80 <- (ONpop2020adult80 - latestfullvaxx)
target75 <- (ONpop2020adult75 - latestfullvaxx)
target70 <- (ONpop2020adult70 - latestfullvaxx)
target65 <- (ONpop2020adult65 - latestfullvaxx)
target60 <- (ONpop2020adult60 - latestfullvaxx)
target55 <- (ONpop2020adult55 - latestfullvaxx)
target50 <- (ONpop2020adult50 - latestfullvaxx)
target45 <- (ONpop2020adult45 - latestfullvaxx)
target40 <- (ONpop2020adult40 - latestfullvaxx)
target35 <- (ONpop2020adult35 - latestfullvaxx)
target30 <- (ONpop2020adult30 - latestfullvaxx)
target25 <- (ONpop2020adult25 - latestfullvaxx)
target20 <- (ONpop2020adult20 - latestfullvaxx)
target15 <- (ONpop2020adult15 - latestfullvaxx)
target10 <- (ONpop2020adult10 - latestfullvaxx)
target05 <- (ONpop2020adult05 - latestfullvaxx)
```


```{r average number a day becomming fully vaccinated, include=FALSE}
# Identify the average number of people a day who become fully vaccinated.
avefullyvaxxday <- mean(individualsfullyvaccinated$change)
```


```{r number of days to meet full target, include = FALSE}
# Calculate the number of days required to meet the share of the adult population vaccinated as a whole number.
daystofull100 <- round(target100/avefullyvaxxday, digits = 0)
daystofull95 <- round(target95/avefullyvaxxday, digits = 0)
daystofull90 <- round(target90/avefullyvaxxday, digits = 0)
daystofull85 <- round(target85/avefullyvaxxday, digits = 0)
daystofull80 <- round(target80/avefullyvaxxday, digits = 0)
daystofull75 <- round(target75/avefullyvaxxday, digits = 0)
daystofull70 <- round(target70/avefullyvaxxday, digits = 0)
daystofull65 <- round(target65/avefullyvaxxday, digits = 0)
daystofull60 <- round(target60/avefullyvaxxday, digits = 0)
daystofull55 <- round(target55/avefullyvaxxday, digits = 0)
daystofull50 <- round(target50/avefullyvaxxday, digits = 0)
daystofull45 <- round(target45/avefullyvaxxday, digits = 0)
daystofull40 <- round(target40/avefullyvaxxday, digits = 0)
daystofull35 <- round(target35/avefullyvaxxday, digits = 0)
daystofull30 <- round(target30/avefullyvaxxday, digits = 0)
daystofull25 <- round(target25/avefullyvaxxday, digits = 0)
daystofull20 <- round(target20/avefullyvaxxday, digits = 0)
daystofull15 <- round(target15/avefullyvaxxday, digits = 0)
daystofull10 <- round(target10/avefullyvaxxday, digits = 0)
daystofull05 <- round(target05/avefullyvaxxday, digits = 0)
```


```{r projected full date, include = FALSE}
# Calculate projected date targets for fully vaccinated will be met from today's date.
date100 <- Sys.Date() + daystofull100
date95 <- Sys.Date() + daystofull95
date90 <- Sys.Date() + daystofull90
date85 <- Sys.Date() + daystofull85
date80 <- Sys.Date() + daystofull80
date75 <- Sys.Date() + daystofull75
date70 <- Sys.Date() + daystofull70
date65 <- Sys.Date() + daystofull65
date60 <- Sys.Date() + daystofull60
date55 <- Sys.Date() + daystofull55
date50 <- Sys.Date() + daystofull50
date45 <- Sys.Date() + daystofull45
date40 <- Sys.Date() + daystofull40
date35 <- Sys.Date() + daystofull35
date30 <- Sys.Date() + daystofull30
date25 <- Sys.Date() + daystofull25
date20 <- Sys.Date() + daystofull20
date15 <- Sys.Date() + daystofull15
date10 <- Sys.Date() + daystofull10
date05 <- Sys.Date() + daystofull05

```

```{r create vaccination targets data frame, echo = FALSE}
# Create data frame for predicted vaccination targets
targets0 <- c("100%", "95%", "90%", "85%", "80%", "75%", "70%", "65%", "60%", "55%", "50%", "45%", "40%", "35%", "30%", "25%", "20%", "15%", "10%", "05%") # column names
targets1 <- c(daystofull100, daystofull95, daystofull90, daystofull85, daystofull80, daystofull75, daystofull70, daystofull65, daystofull60, daystofull55, daystofull50, daystofull45, daystofull40, daystofull35, daystofull30, daystofull25, daystofull20, daystofull15, daystofull10, daystofull05) # projected number of days to meet target
targets2 <- c(date100, date95, date90, date85, date80, date75, date70 ,date65, date60, date55, date50, date45, date40, date35, date30, date25, date20, date15, date10, date05) # projected date target will be met

# Bind data frame together
targets <- data.frame(targets0, targets1, targets2)
names(targets) <- c("Population vaccination target", "Projected days to full vaccination", "Projected date for full vaccination") # rename columns
knitr::kable(targets, align = "c")
```

## Citations
### Data
Ontario COVID-19 Vaccine Data https://data.ontario.ca/dataset/covid-19-vaccine-data-in-ontario/resource/8a89caa9-511c-4568-af89-7f2174b4378c?view_id=9e42f55b-723f-46dd-b0d9-643670e01fed published under Open Government Licence – Ontario version 1.0 (https://www.ontario.ca/page/open-government-licence-ontario)

Statistics Canada. Table 17-10-0005-01  Population estimates on July 1st, by age and sex. https://doi.org/10.25318/1710000501-eng published under Statistics Canada Open Licence https://www.statcan.gc.ca/eng/reference/licence

### Code
Horton, L. (2021). Ontario Covid Vaccinations. *GitHub repository*, https://github.com/laurencehorton/ontario-covid-vaccinations published under Creative Commons Zero v1.0 Universal https://github.com/laurencehorton/ontario-covid-vaccinations/blob/main/LICENSE 

```{r citations, include = FALSE}
toBibtex(citation("base"))
toBibtex(citation("knitr"))
toBibtex(citation("lubridate"))
toBibtex(citation("rjson"))
toBibtex(citation("tidyverse"))
toBibtex(citation("rmarkdown"))